---
title: "Air Quality :VAR MODEL"
output:
  word_document: default
  html_document:
    df_print: paged
---
## Members
- ANDREW KARANJA
- SUHAYMA ALI HARITH
- JULIET WANJIRU 
- HAPPY KULOLA
- ALEX WAWERU  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

```{r libraries, include=FALSE}
library(summarytools)
library(tidyverse)
library(VIM)
library(lubridate)
library(tseries)
library(lmtest)
library(fUnitRoots)
library(forecast)
library(imputeTS)
library(vars)
library(ggplot2)
```

# Importing Data {.tabset}
## Region 1
```{r importing data}
Region_1=read.csv("Region A.csv")
head(Region_1,4)
#Selecting the data from the first day we had an active reading in region 1
Region_1=Region_1[281:length(Region_1$Date),]
```

## Region 2
```{r}
Region_2=read.csv("Region B.csv")
head(Region_2,4)
#Selecting the data from the first day we had an active reading in region 2
Region_2=Region_2[66:length(Region_2$Date),]
```

>The days containing missing values indicate that that particlular day no data was collected in that region

# EDA {.tabset}
## Region 1
```{r summary of the data}
summarytools::view(dfSummary(Region_1,graph.col = F))
```

## Region 2
```{r}
summarytools::view(dfSummary(Region_2,graph.col = F))
```

# Data Manipulation {.tabset}
Creating new variables extracted from date and reshaping the data 
```{r}
#Creating a data frame
Data=rbind(data.frame(Region_1,Region=rep("A",length(Region_1$Date))),
           data.frame(Region_2,Region=rep("B",length(Region_2$Date))))

#Creating a variable day
Data$Day=lubridate::wday(Data$Date,label = T)

#Creating a variable month
Data$Month=lubridate::month(Data$Date,label = T)

#Creating a variable year
Data$Year=factor(lubridate::year(Data$Date))

#Reshaping the data
Data=Data %>% gather('Value_Type','Value',
                             -c('Date','Day','Month','Year','Region'))

```

## Removing Outliers
```{r}
R=unique(Data$Region)
Vt=unique(Data$Value_Type)

for(i in R){
  for (j in Vt) {
   Outlier=boxplot.stats(Data$Value[which(Data$Region==paste(i) & 
                                 Data$Value_Type==paste(j))])$out
   Data = Data %>% filter(!(Value %in% Outlier))
  }
}
```

# Distribution {.tabset}
## Temperature 
```{r}
Data %>% filter(Value_Type=="Temp") %>% 
  ggplot(aes(Value)) + 
  geom_density(col="red") +
  theme_dark() +
  ggtitle("Temperatures Distribution")
  
```

## Humidity 
```{r}
Data %>% filter(Value_Type=="Humid") %>% 
  ggplot(aes(Value)) + 
  geom_density(col="blue") +
  theme_dark()+
  ggtitle("Humidity Distribution")
  
```

## Particulate matter 1
```{r}
Data %>% filter(Value_Type=="p1") %>% 
  ggplot(aes(Value)) + 
  geom_density(col="green") +
  theme_dark()+
  ggtitle("Particulate matter 1 Distribution")
  
```

## Particulate matter 1
```{r}
Data %>% filter(Value_Type=="p2") %>% 
  ggplot(aes(Value)) + 
  geom_density(col="yellow") +
  theme_dark()+
  ggtitle("Particulate matter 2 Distribution")
  
```

# Visualization {.tabset}
## Yearly Analysis 
```{r Year}
Data %>% #filter(Value_Type=="Temp") %>% 
  ggplot(aes(x=Year,y=Value,fill=Value_Type)) +
  geom_boxplot() +
  facet_wrap(~Region)
```

## Monthly Analysis
```{r Month}
Data %>% #filter(Value_Type=="Temp") %>%   
  ggplot(aes(x=Month,y=Value,fill=Value_Type)) +
  geom_boxplot() +
  facet_wrap(~Region) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=1))
```

## Daily Analysis
```{r Day}
Data %>% ggplot(aes(x=Day,y=Value,fill=Value_Type)) +
  geom_boxplot() +
  facet_wrap(~Region)
```
## Table
```{r}
Tab=Data %>% group_by(Value_Type,Region,Month) %>% 
         summarise(Value=median(Value,na.rm=T),.groups = "keep") %>%
  spread(Value_Type,Value)

Tab %>% filter(Region=="A") %>% ggplot(aes(x=Month)) + 
  geom_point(aes(y=Temp,col="Temp")) +
  geom_point(aes(y=Humid,col="Humid")) +
  geom_point(aes(y=p1,col="P2.5")) +
  geom_point(aes(y=p2,col="p10")) +
  ggtitle("Region A Median Recorded Value per Month")

Tab %>% filter(Region=="B") %>% ggplot(aes(x=Month)) + 
  geom_point(aes(y=Temp,col="Temp")) +
  geom_point(aes(y=Humid,col="Humid")) +
  geom_point(aes(y=p1,col="P2.5")) +
  geom_point(aes(y=p2,col="p10"))+
  ggtitle("Region A Median Recorded Value per Month")
```

---

# Vector Auto-Regression (VAR)

## Converting data to a time series
```{r}
Region_1ts=ts(Region_1[-1],frequency = 365,
              start = c(2018,282))

Region_2ts=ts(Region_2[-1],frequency = 365,
              start = c(2018,67))
```

## Replacing outliers present in the two regions time series data
```{r outlier}
#Detecting outliers and replacing them
for(i in 1:4){
  uni=Region_1ts[,i]
  for(j in 1:length(tsoutliers(uni)$index)){
    Region_1ts[tsoutliers(uni)$index[j],i]=tsoutliers(uni)$replacements[j]
  }
}

for(i in 1:4){
  uni=Region_2ts[,i]
  for(j in 1:length(tsoutliers(uni)$index)){
    Region_2ts[tsoutliers(uni)$index[j],i]=tsoutliers(uni)$replacements[j]
  }
}
```

## Handling missing values {.tabset}
### Region 1
```{r }
# Visualization
any(is.na(Region_1))

colSums(is.na(Region_1))

aggr(Region_1, col=c('navyblue','yellow'),
                    numbers=TRUE, sortVars=TRUE,
                    labels=names("Region 1"), cex.axis=.7,
                    gap=3, ylab=c("Missing data","Pattern"))


plot(Region_1ts, main="Region 1 Before Imputation")

# Missing values imputation
Region_1imputed=na_seadec(Region_1ts)

#After Imputation
plot(Region_1imputed, main="Region 1 After Imputation")
```

### Region 2
```{r }
# Visualization
any(is.na(Region_2))

colSums(is.na(Region_2))

aggr(Region_2, col=c('navyblue','yellow'),
                    numbers=TRUE, sortVars=TRUE,
                    labels=names("Region 2"), cex.axis=.7,
                    gap=3, ylab=c("Missing data","Pattern"))
# Before imputation
plot(Region_2ts, main="Region 2 Before Imputation")

# Missing values imputation
Region_2imputed=na_seadec(Region_2ts)

#After Imputation
plot(Region_2imputed, main="Region 2 After Imputation")
```


## Granger’s Causality Test
```{r}
#Test of causality between Temp and humidity
grangertest(Region_1imputed[,1],Region_1imputed[,2],order = 2)$`Pr(>F)`
grangertest(Region_2imputed[,1],Region_2imputed[,2],order = 2)$`Pr(>F)`

#Test of causality between P1 and P2
grangertest(Region_1imputed[,4],Region_1imputed[,3],order = 2)$`Pr(>F)`
grangertest(Region_2imputed[,4],Region_2imputed[,3],order = 2)$`Pr(>F)`

#Test of causality between humid and P1
grangertest(Region_1imputed[,2],Region_1imputed[,3],order = 2)$`Pr(>F)`
grangertest(Region_2imputed[,2],Region_2imputed[,3],order = 2)$`Pr(>F)`
```


## Test Stationarity {.tabset}

### Region 1
Augmented Dickey-Fuller test for the null hypothesis that x is non-stationary.
```{r Dicky Fuller Test,decomposition and detrending 1 }
lapply(Region_1imputed,adfTest,lags=1,type="nc")

## Decomposition
autoplot(decompose(Region_1imputed[,1]),main="Temperature")
autoplot(decompose(Region_1imputed[,2]),main="Humidity")
autoplot(decompose(Region_1imputed[,3]),main="P1")
autoplot(decompose(Region_1imputed[,4]),main="P2")

## Differencing
Region1_diff=diff(Region_1imputed,lag = 1)

autoplot(decompose(Region1_diff[,1]),main="Temperature")
autoplot(decompose(Region1_diff[,2]),main="Humidity")
autoplot(decompose(Region1_diff[,3]),main="P1")
autoplot(decompose(Region1_diff[,4]),main="P2")

acf(Region1_diff)
```

Since the p-values > 0.05, we conclude that there is no enough evidence to reject the Null hypothesis, meaning that the time series is non-stationary.

### Region 2
```{r Dicky Fuller Test,decomposition and detrending 2}
lapply(Region_2imputed,adfTest,lags=1, type="nc")

## Decomposition
autoplot(decompose(Region_2imputed[,1]),main="Temperature")
autoplot(decompose(Region_2imputed[,2]),main="Humidity")
autoplot(decompose(Region_2imputed[,3]),main="P1")
autoplot(decompose(Region_2imputed[,4]),main="P2")

## Differencing
Region2_diff=diff(Region_2imputed,lag = 1)

autoplot(decompose(Region2_diff[,1]),main="Temperature")
autoplot(decompose(Region2_diff[,2]),main="Humidity")
autoplot(decompose(Region2_diff[,3]),main="P1")
autoplot(decompose(Region2_diff[,4]),main="P2")

acf(Region2_diff)
```

Since the p-values < 0.05, we conclude that there is enough evidence to reject the Null hypothesis, meaning that the time series is stationary.

## Model {.tabset}
### Region 1
```{r Model1}
Model_1=VAR(Region_1imputed,p=1,type = "const", season = 12)
summary(Model_1)

#residual diagnostics
serial.test(Model_1)
arch.test(Model_1)

fcast = predict(Model_1, n.ahead = 5)
fcast

var_forecast = forecast(Model_1, h=30)
autoplot(var_forecast, main="VAR Forecast", xlab="Date")
```

### Region 2
```{r Model2}
Model_2=VAR(Region_2imputed,p=1,type = "const", season = 12)
summary(Model_2) 

#residual diagnostics
serial.test(Model_2)
arch.test(Model_2)

fcast = predict(Model_2, n.ahead = 5)
fcast

var_forecast = forecast(Model_2, h=30)
autoplot(var_forecast, main="VAR Forecast", xlab="Date")

```
